---
title: "Barrier_Exercise_MWTL"
author: "Alicia Hamer"
format: html
editor: visual
execute: 
  echo: true
  eval: true
  cache: false
---

# To Do:

- Empty stations need to be filled
- Grevelingen stations need to be removed

# Case study: shellfish in two tidal inlets separated by a barrier (MWTL survey)

The Oosterschelde and Westerschelde are two adjacent tidal systems in the Dutch Delta. Although they are spatially close, they are physically separated by land so spatial correlation in benthos density should *not* flow freely across the peninsula, even where two sampling points are close in straight-line distance. This case study tests whether encoding that barrier in the spatial mesh improves a species distribution model.

This version uses the **MWTL macrozoobenthos survey** (MWTL_MZB) instead of the WOT shellfish survey. The MWTL survey is not restricted to mudflats, so it should sample a broader range of the study area. We model **density (N_m2)** rather than raw counts, using a Tweedie family.

We fit two otherwise-identical `sdmTMB` models: one on a regular SPDE mesh and one on a barrier mesh, and compare.

# 0. Information on Survey:

## Settings

```{r settings}
#    Settings

crs_chosen     =    4326   #WGS84
Species_chosen = c( "Peringia ulvae","Corophium volutator",       #littoral species
                   "Pygospio elegans","Heteromastus filiformis"   #sublittoral species
                   )[1]   
Season_chosen  = "nj"                      # focus on najaar (autumn)
Year_end       = 2009                      #sampling design changed in 2009 

```

## Load libraries

```{r libs}
#| message: false
#| warning: false
#| cache: false

x <- c("ggplot2", "tidyverse", "vegan", "sf", "raster", "terra", "tidyterra","glmmTMB",
       "dplyr", "tidyr", "sdmTMB", "sdmTMBextra", "DHARMa", "patchwork")
invisible(lapply(x, require, character.only = TRUE))

#Set Theme
theme_set(theme_bw())

#Directories
path          <- getwd()
pathfigures   <- paste0(path,"/Fig")

```

## 0.1 Read Survey data

Read the Oosterschelde and Westerschelde MWTL survey files separately, then merge.

Columnnames key: - `Valid_name` = accepted species name - `N_m2` = density per m2 - `N` = raw count - `Oppervlakte_m2` = sampled area - `Tuig` = gear - `Seizoen` = season (`vj` / `nj`) - `StationID` = station identifier - `Dvd` = emersion time (extracted, not measured during survey) - `EvMax` = Maximum flow velocity (extracted, not measured during survey)

```{r Read In Survey}
#| message: false
#| warning: false

#Oosterschelde
df_MWTL_OS <- read_csv(paste0(path, "/data/SpeciesData/BenthosSurvey/mwtl_os.csv")) %>%   # <<< set filename
  dplyr::mutate(StationID = as.character(StationID)) %>%
  dplyr::mutate(Systeem = "Oosterschelde")

#Westerschelde
#cutting off the Waddensea samples (same survey but not part of the case study)
df_MWTL_WS <- read_csv(paste0(path, "/data/SpeciesData/BenthosSurvey/mwtl_ws.csv")) %>%   # <<< set filename
  dplyr::filter(Y < 52.0) %>%                       # Y is Latitude in this file
  dplyr::mutate(StationID = as.character(StationID)) %>%
  dplyr::mutate(Systeem = "Westerschelde")

#Merge the two systems
df_MWTL <- dplyr::bind_rows(df_MWTL_OS, df_MWTL_WS)

# Rename to neutral internal names so the rest of the script is stable.
# (Longitude/Latitude kept as explicit columns; X/Y get reused later for UTM.)
df_MWTL <- df_MWTL %>%
  dplyr::rename(
    Longitude = X,
    Latitude  = Y,
    dvd       = Dvd,
    evmax     = Evmax
  )

#Merge sampling gears into coarse classes (adjust to the gear codes actually present)
df_MWTL <- df_MWTL %>%
  mutate(
    Tuig = dplyr::case_when(
      Tuig %in% c("STEEKBS", "Steekbuis", "steekbuis (pvc-buis)") ~ "Steekbuis",
      Tuig %in% c("BOXCORE", "Boxcorer")                          ~ "Boxcore",
      TRUE                                                        ~ "Overig"
    ),
    Tuig = factor(Tuig)
  ) %>%
  dplyr::filter(Projecten == "MWTL_MZB") %>%     # keep only the MWTL macrozoobenthos project
  dplyr::filter(Seizoen == Season_chosen)         # focus on nj

#End at selected year (in case you want it up until the survey design change in 2009 (set in settings in first section))
df_MWTL <- df_MWTL %>% dplyr::filter(Year <= Year_end)

#Turn it into a spatial object
df_MWTL_sf  <- st_as_sf(df_MWTL, coords = c("Longitude","Latitude"),
                        crs = crs_chosen, remove = FALSE)


```

## 0.2 Read Env data

The main covariate is emersion time (*DVD -\> Droogvalduur (Dutch)*), supplied as rasters per system. We reproject both to WGS84, merge them, and load the land barrier shapefile (cropped to the study extent) for use in plotting and in the barrier mesh.

```{r Read in Environmental Data}
#| message: false
#| warning: false

#Emersion time       (Unit is percentage % of time emerged)
dvd_OS       <- terra::rast(paste0(path, "/Data/EnvironData/Rasters_EmersionTimes", "/edvd_os_2021.tif"))
dvd_WS       <- terra::rast(paste0(path, "/Data/EnvironData/Rasters_EmersionTimes", "/edvd_ws_2022.tif"))

#Flow Velocity (Unit is m/sec)
#convert from rd new to wgs84
dvd <- terra::merge(terra::sprc(dvd_OS, dvd_WS))
dvd <- terra::project(dvd, "EPSG:4326", method = "bilinear")

# Create breaks
breaks <- c(0,5,10,15,20,25,30,35,40,45,50,60,70,80,90,100)


#Flow Velocity      (Unit is m/sec)
evmax_OS       <- terra::rast(paste0(path, "/Data/EnvironData/Rasters_EvMax", "/evmax_os_2021.tif"))
evmax_OS <- terra::resample(evmax_OS, dvd_OS, method = "bilinear")  # align grids
evmax_OS <- terra::mask(evmax_OS, dvd_OS)

evmax_WS <- terra::rast(paste0(path, "/Data/EnvironData/Rasters_EvMax", "/evmax_ws_2022.tif"))
evmax_WS <- terra::resample(evmax_WS, dvd_WS, method = "bilinear") 
evmax_WS <- terra::mask(evmax_WS, dvd_WS)

#convert from rd new to wgs84
ev_max <- terra::merge(terra::sprc(evmax_OS, evmax_WS))
ev_max <- terra::project(ev_max, "EPSG:4326", method = "bilinear")
#very important step I forgot! 
ev_max <- ev_max/100 

#Barrier_Shapefile   -> this marks the tidal flats as barriers and that is not what we want
Barrier            <- st_read(paste0(path, "/Data/EnvironData/Shapefile_Boundary", "/Nederland.shp")) 
st_crs(Barrier)    <- 32631  
Barrier            <- st_transform(Barrier, crs_chosen)

#Cut it
Barrier           <- st_geometry(Barrier)
Barrier           <- st_make_valid(Barrier)
Barrier           <- st_crop(Barrier, c(xmin = 3.35, ymin = 51.25, xmax = 4.30, ymax = 51.75))

#Construct barrier from Emersion time (dvd) grid layer
land_rast <- terra::ifel(is.na(dvd), 1, NA)        # 1 where dvd is NA (= land), NA elsewhere
land      <- terra::as.polygons(land_rast)          # polygon of the land
land      <- sf::st_as_sf(land)                      # to sf, inherits dvd's CRS (4326)
land      <- sf::st_make_valid(land)
#plot(sf::st_geometry(land), col = "tomato", border = "grey40")

Barrier   <- land


rm(list = setdiff(ls(), c("df_MWTL", "df_MWTL_sf",                         #data
                          "Barrier", "dvd", "ev_max",                      #spatial data
                          "path", "pathfigures",                           #paths
                          "Species_chosen", "Surveyx", "breaks"            #settings
                          )))

```

# 1. Area description

## 1.1 Area description - Spatially

The westerschelde is more dynamic (larger flow velocities) and has a salinity gradient, the oosterschelde is less dynamic and no salinity gradient. Both are tidal systems.

The map below shows the emersion-time surface, the land barrier separating the two inlets, and the survey stations coloured by sampling gear.

```{r plot descriptive}
#| message: false
#| echo: false

p1 <- ggplot() +
  tidyterra::geom_spatraster(data = dvd) +
  scale_fill_binned(palette =  c("#08306B","#08519C","#2171B5","#4292C6","#6BAED6",
                                 "#9ECAE1","#C6DBEF","#DEEBF7","#F7FBFF",
                                 "lightyellow","lightyellow1","lightyellow2",
                                 "lightyellow3","lightyellow4", "yellow4",NA),
                    breaks = breaks,
                    na.value = "white") +
  geom_sf(data = Barrier, fill = "grey85", colour = "grey40", linewidth = 0.2) +
  geom_sf(data = df_MWTL_sf, aes(colour = Tuig), size = 0.6, alpha = 0.6) +
  coord_sf(xlim = c(3.35, 4.30), ylim = c(51.25, 51.75), expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_bw()

print(p1)

```

## 1.2 Area description - Table

```{r Table}
#| message: false
#| echo: false

# One row per sampling event. In MWTL long format there is no replicate.id,
# so an event is one Station x Date (x gear). Build an event key.
df_MWTL <- df_MWTL %>%
  dplyr::mutate(event_id = paste(StationID, Date, Tuig, sep = "_"))

events_tagged <- df_MWTL %>%
  dplyr::distinct(event_id, .keep_all = TRUE)

# Samples per system per year
samples_per_system <- events_tagged %>%
  dplyr::count(Year, Systeem) %>%
  tidyr::pivot_wider(names_from = Systeem, values_from = n, values_fill = 0) %>%
  dplyr::arrange(Year)

knitr::kable(samples_per_system,
             caption = "Number of survey samples per system per year")

# Sampled environment (dvd, evmax) per system
env_per_system <- events_tagged %>%
  dplyr::filter(!is.na(dvd), !is.na(evmax)) %>%
  dplyr::group_by(Systeem) %>%
  dplyr::summarise(
    n          = dplyr::n(),
    dvd_mean   = mean(dvd),   dvd_sd   = sd(dvd),
    evmax_mean = mean(evmax), evmax_sd = sd(evmax),
    .groups = "drop"
  )

knitr::kable(env_per_system, digits = 2,
             caption = "Sampled emersion time (dvd) and max flow velocity (evmax) by system")
```

## 1.3 Look at species' distribution over years

See where species is located per year and if this changes

```{r Species}
#| message: false
#| echo: false
#| fig-height: 12
#| fig-width: 9

#look at data
#up until 2009 only? This is when they changed survey design so maybe that helps
df_plot <- df_MWTL_sf %>% dplyr::filter(Year < 2010)

ggplot(df_plot, aes(Longitude, Latitude, size = N_m2, colour = log(N_m2 + 1))) +
  geom_point() +
  facet_wrap(~Year, ncol = 3) +
  ggtitle(paste(Species_chosen))


```

# 2. Model

Prepare data for model

```{r Data processing}


# One row per event, carrying the covariates/metadata
events <- df_MWTL %>%
  group_by(event_id) %>%
  dplyr::summarise(
    Year      = first(Year),
    Longitude = first(Longitude),
    Latitude  = first(Latitude),
    dvd       = first(dvd),
    evmax     = first(evmax),
    Tuig      = first(Tuig),
    Seizoen   = first(Seizoen),
    Systeem   = first(Systeem),
    .groups = "drop"
  )

# Density of the chosen species per event (from N_m2)
sp_density <- df_MWTL %>%
  dplyr::filter(Valid_name == Species_chosen) %>%
  group_by(event_id) %>%
  dplyr::summarise(density = sum(N_m2, na.rm = TRUE), .groups = "drop")

dat <- events %>%
  left_join(sp_density, by = "event_id") %>%
  mutate(density = ifelse(is.na(density), 0, density))   # true absences -> 0 density

cat("Events with species present:", sum(dat$density > 0),
    " | absences:", sum(dat$density == 0), "\n")

dat <- dat %>%
  filter(!is.na(dvd), !is.na(evmax),
         !is.na(Longitude), !is.na(Latitude)) %>%
  mutate(
    dvd_s    = as.numeric(scale(dvd)),
    evmax_s  = as.numeric(scale(evmax)),
    gear     = factor(Tuig),
    fYear    = factor(Year)
  )

## Project lon/lat -> metric CRS (UTM 31N) so mesh & range are in km.
dat <- add_utm_columns(dat, ll_names = c("Longitude", "Latitude"),
                       utm_crs = 32631)   # adds X, Y (km)



```

## 2.1 Constructing a Barrier Mesh

Build a regular SPDE mesh first, then derive a barrier mesh from it by passing the land polygon to `add_barrier_mesh()`. The `range_fraction` shrinks the spatial range across the barrier (lower = stronger barrier), so correlation no longer leaks across land.

```{r mesh}

#regular mesh
mesh <- sdmTMB::make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 0.75)
plot(mesh); cat("Mesh vertices:", mesh$mesh$n, "\n")
plot(mesh$mesh, main = "Regular mesh", asp = 1)

Barrier <- sf::st_sf(geometry = sf::st_geometry(Barrier))
Barrier_km <- sf::st_transform(Barrier, 32631)          # metres
sf::st_geometry(Barrier_km) <- sf::st_geometry(Barrier_km) * 1e-3   # -> km to match mesh X/Y
sf::st_crs(Barrier_km) <- NA

#barrier mesh
barrier_mesh <- sdmTMBextra::add_barrier_mesh(
  spde_obj    = mesh,
  barrier_sf  = Barrier_km,
  range_fraction = 0.2,
  proj_scaling = 1,
  plot = TRUE
)


#plot(barrier_mesh$mesh, main = "Barrier mesh")

```

## 2.2 Fitting the models

```{r Model}

fit_plain <- sdmTMB(
  density ~ poly(dvd_s, 2) + poly(evmax_s, 2) + fYear,
  data   = dat,
  mesh   = mesh,
  family = tweedie(link = "log"),
  spatial = "on"
)


fit_barrier <- sdmTMB(
  density ~ poly(dvd_s, 2) + poly(evmax_s, 2) + fYear,
  data   = dat,
  mesh   = barrier_mesh,
  family = tweedie(link = "log"),
  spatial = "on"
)

```

## 2.3 Setting up Prediction grid

```{r Prediction-grid}

# Coarsen the emersion raster a little so prediction is quick,

# --- dvd grid
dvd_coarse     <- terra::aggregate(dvd, fact = 4, fun = "mean", na.rm = TRUE)
grid           <- terra::as.data.frame(dvd_coarse, xy = TRUE, na.rm = TRUE)
names(grid)[3] <- "dvd"

# --- pull evmax 
evmax_coarse <- terra::resample(ev_max, dvd_coarse, method = "bilinear")
grid$evmax   <- terra::extract(evmax_coarse,
                               grid[, c("x", "y")])[, 2] 

# drop cells where evmax is missing and filter out subtidal area for prediction grid -> not sure is this is best practice
grid <- grid[!is.na(grid$evmax), ]
#grid <- grid[grid$dvd > 0 & grid$dvd < 97, ]  #not sure if neccesary

# --- scale BOTH using the fitting means/sds ---
dvd_mean   <- mean(dat$dvd,   na.rm = TRUE); dvd_sd   <- sd(dat$dvd,   na.rm = TRUE)
evmax_mean <- mean(dat$evmax, na.rm = TRUE); evmax_sd <- sd(dat$evmax, na.rm = TRUE)

grid <- grid %>%
  dplyr::rename(Longitude = x, Latitude = y) %>%
  dplyr::mutate(
    dvd_s   = (dvd   - dvd_mean)   / dvd_sd,
    evmax_s = (evmax - evmax_mean) / evmax_sd
  )

# gear still held at reference level 
grid$gear <- factor(levels(dat$gear)[1], levels = levels(dat$gear))

prediction_year <- 2009
grid$fYear <- factor(prediction_year, levels = levels(dat$fYear))

grid <- add_utm_columns(grid, ll_names = c("Longitude", "Latitude"),
                        utm_crs = 32631)

cat("Prediction cells:", nrow(grid), "\n")

```

# 3. Comparison

## 3.1 Performance

```{r Model Performance}
#| message: false

sanity(fit_plain)
sanity(fit_barrier)

print(fit_plain)
print(fit_barrier)

# AIC 
AIC(fit_plain, fit_barrier)  

#DHARMa residuals  
s1 <- simulate(fit_plain,   nsim = 300, type = "mle-mvn")
s2 <- simulate(fit_barrier, nsim = 300, type = "mle-mvn")

r1 <- dharma_residuals(s1, fit_plain,   return_DHARMa = TRUE)
r2 <- dharma_residuals(s2, fit_barrier, return_DHARMa = TRUE)

#DHARMa::testResiduals(r1)
#DHARMa::testResiduals(r2)
#DHARMa::testSpatialAutocorrelation(r1, x = dat$X, y = dat$Y)
#DHARMa::testSpatialAutocorrelation(r2, x = dat$X, y = dat$Y)

#S
tidy(fit_plain, "ran_pars", conf.int = TRUE)
tidy(fit_barrier, "ran_pars", conf.int = TRUE)

```

## 3.2 Prediction

```{r Prediction-predict}

pred_plain   <- predict(fit_plain,   newdata = grid)
pred_barrier <- predict(fit_barrier, newdata = grid)

pred_plain$density   <- exp(pred_plain$est)
pred_barrier$density <- exp(pred_barrier$est)

```

## 3.3 Conditional effects

Can't get these to work when rendering :(

```{Conditional effects}

#these can take a bit to load
visreg::visreg(fit_plain, "dvd_s")
visreg::visreg(fit_plain, "evmax_s")

#ggeffects::ggeffect(fit_plain, "dvd_s [0:10 by=1]") %>%  plot()

print(plot(PartDvd))
#print(PartEvmax)



```

## 3.4 Spatial prediction plots

```{r Prediction-maps, fig.width=11, fig.height=5}

zlim <- range(c(pred_plain$est, pred_barrier$est), na.rm = TRUE)

base_map <- function(df, title) {
  ggplot(df, aes(Longitude, Latitude, fill = est)) +
    geom_sf(data = Barrier, fill = "grey85", colour = "grey40",
            linewidth = 0.2, inherit.aes = FALSE) +
    geom_raster() +
    scale_fill_viridis_c(name = "log expected\ncount",
                         limits = zlim) +             
    coord_sf(xlim = c(3.35, 4.30), ylim = c(51.25, 51.75), expand = FALSE) +
    labs(title = title, x = NULL, y = NULL) +
    theme_bw()
}

p_plain   <- base_map(pred_plain,   paste0(Species_chosen, " Regular mesh"))
p_barrier <- base_map(pred_barrier, paste0(Species_chosen, " Barrier mesh"))

library(patchwork)
p_plain + p_barrier


print(fit_plain)
print(fit_barrier)

#nice to hear our sampling effore is good enough to .. although MWTL is not every year so sruely if you want to get temproal shifts in distribution it could cause issues? 

```

# Reccomendations

The DFS data could be added to explore if mesh/barrier mesh shows a difference in fish species, maybe moving species could react differently?

Or would a fish survey run against the same problems?
