---
title: "Barrier_Exercise"
author: "Alicia Hamer"
format: 
  gfm:
    output-file: "README.md"
editor: visual
execute: 
  echo: true
  eval: true
  cache: false
---

# Case study: shellfish in two tidal inlets separated by a barrier

The Oosterschelde and Westerschelde are two adjacent tidal systems in the Dutch Delta. Although they are spatially close, they are physically separated by land so spatial correlation in shellfish density should *not* flow freely across the peninsula, even where two sampling points are close in straight-line distance. This case study tests whether encoding that barrier in the spatial mesh improves a species distribution model for the Common Cockle (*Cerastoderma edule*).

The data come from the annual intertidal **Kokkelsurvey**, conducted each spring by Wageningen Marine Research. Depending on the tide at the moment of sampling, either a *kokkelschepje* or a *steekbuis* is used to take the sample.

We fit two otherwise-identical `sdmTMB` models: one on a regular SPDE mesh and one on a barrier mesh, and compare.

# 0. Information on Survey:

Monster Identifier: replicate.id Monster surface area:

## Settings

Here we can choose the survey and the species, for this case study the Kokkelsurvey in the spring and the Common Cockle (Cerastoderma edule) might be most interesting. (*Ruditapes* is an invasive species, so probably less suitable).

```{r settings}
#    Settings

Surveyx        = c("KokkelSurveyVoorjaar", "KokkelSurveyNajaar", "Oestersurvey")[1] #I removed other surveys so this does not do anything right now
crs_chosen     =    4326   #WGS84
Species_chosen = c("Cerastoderma edule", "Macoma balthica", "Ruditapes philippinarum", "Abra tenuis")[1]

```

## Load libraries

```{r libs}
#| message: false
#| warning: false
#| cache: false

x <- c("ggplot2", "tidyverse", "vegan", "sf", "raster", "terra", "tidyterra","glmmTMB",
       "dplyr", "tidyr", "sdmTMB", "DHARMa")                  
invisible(lapply(x, require, character.only = TRUE))

#Set Theme
theme_set(theme_bw())                           

#Directories
path          <- getwd()
pathfigures   <- paste0(path,"/Fig")

```

## 0.1 Read Survey data

Read the Oosterschelde and Westerschelde survey files separately, drop the Wadden Sea stations from the Westerschelde file (they share a survey but are outside this case study), and merge the two systems into one table.

```{r Read In Survey}
#| message: false
#| warning: false

#Oosterschelde
df_WOT_OS <- read_csv(paste0(path, "/data/SpeciesData/ShelfishSurvey/df_WOT_os.csv")) %>% 
  dplyr::mutate(locatie.nr = as.character(locatie.nr)) %>% 
  dplyr::mutate(Systeem = "Oosterschelde")

#Westerschelde 
#cutting of the Waddensea samples (They belong to the same survey as the Westerschelde but aren't part of the case study)
df_WOT_WS <- read_csv(paste0(path, "/data/SpeciesData/ShelfishSurvey/df_WOT_ws.csv")) %>% 
  dplyr::filter(Latitude < 52.0) %>% 
  dplyr::mutate(locatie.nr = as.character(locatie.nr)) %>% 
  dplyr::mutate(Systeem = "Westerschelde")
  
#Merge the two systems
df_WOT <- dplyr::bind_rows(df_WOT_OS, df_WOT_WS)

#Remove size/age classes (so there are not multiple rows per unique ID)
# this was already done

#Select Survey and Merge Sampling methos
df_WOT <- df_WOT %>% mutate(
  Tuig = dplyr::case_when(
    Tuig %in% c("Kokkelschepje 1 Regulus",
                "Kokkelschepje 2 Luctor",
                "Kokkelschepje 3 reserve",
                "kokkelschuifje")              ~ "Kokkelschepje",
    Tuig %in% c("steekbuis (pvc-buis)")        ~ "Steekbuis",
    TRUE                                       ~ "Overig"   # ring, OesterHapper, etc.
  ), Tuig = factor(Tuig)) %>%
  dplyr::filter(project.id %in% c(4,5)) %>%   # these numbers are the survey identifiers
  dplyr::filter(seizoen == "vj")              #only spring as this is the long term survey (vj stands for voorjaar)


#Turn it into a spatial object
df_WOT_sf  <- st_as_sf(df_WOT, coords = c("Longitude","Latitude"), crs = crs_chosen, remove = FALSE)

```

## 0.2 Read Env data

The main covariate is emersion time (*DVD -\> Droogvalduur (Dutch)*), supplied as rasters per system. We reproject both to WGS84, merge them, and load the land barrier shapefile (cropped to the study extent) for use in plotting and in the barrier mesh.

```{r Read in Environmental Data}
#| message: false
#| warning: false

#Emersion time       (Unit is percentage % of time emerged)
dvd_OS       <- terra::rast(paste0(path, "/Data/EnvironData/Rasters_EmersionTimes", "/edvd_os_2021.tif"))
dvd_WS       <- terra::rast(paste0(path, "/Data/EnvironData/Rasters_EmersionTimes", "/edvd_ws_2022.tif"))

#convert from rd new to wgs84
dvd <- terra::merge(terra::sprc(dvd_OS, dvd_WS))
dvd <- terra::project(dvd, "EPSG:4326", method = "bilinear")

# Create breaks
breaks <- c(0,5,10,15,20,25,30,35,40,45,50,60,70,80,90,100)


#Flow Velocity      (Unit is m/sec)
evmax_OS       <- terra::rast(paste0(path, "/Data/EnvironData/Rasters_EvMax", "/evmax_os_2021.tif"))
evmax_OS <- terra::resample(evmax_OS, dvd_OS, method = "bilinear")  # align grids
evmax_OS <- terra::mask(evmax_OS, dvd_OS)

evmax_WS <- terra::rast(paste0(path, "/Data/EnvironData/Rasters_EvMax", "/evmax_ws_2022.tif"))
evmax_WS <- terra::resample(evmax_WS, dvd_WS, method = "bilinear") 
evmax_WS <- terra::mask(evmax_WS, dvd_WS)

#convert from rd new to wgs84
ev_max <- terra::merge(terra::sprc(evmax_OS, evmax_WS))
ev_max <- terra::project(ev_max, "EPSG:4326", method = "bilinear")
#very important step I forgot! 
ev_max <- ev_max/100 

#Barrier_Shapefile   -> this marks the tidal flats as barriers and that is not what we want
Barrier            <- st_read(paste0(path, "/Data/EnvironData/Shapefile_Boundary", "/Nederland.shp")) 
st_crs(Barrier)    <- 32631  
Barrier            <- st_transform(Barrier, crs_chosen)

#Cut it
Barrier           <- st_geometry(Barrier)
Barrier           <- st_make_valid(Barrier)
Barrier           <- st_crop(Barrier, c(xmin = 3.35, ymin = 51.25, xmax = 4.30, ymax = 51.75))

#Construct barrier from Emersion time (dvd) grid layer
land_rast <- terra::ifel(is.na(dvd), 1, NA)        # 1 where dvd is NA (= land), NA elsewhere
land      <- terra::as.polygons(land_rast)          # polygon of the land
land      <- sf::st_as_sf(land)                      # to sf, inherits dvd's CRS (4326)
land      <- sf::st_make_valid(land)
#plot(sf::st_geometry(land), col = "tomato", border = "grey40")

Barrier   <- land


rm(list = setdiff(ls(), c("df_WOT", "df_WOT_sf", "MB_perLocation",         #data
                          "Barrier", "dvd", "ev_max",                      #spatial data
                          "path", "pathfigures",                           #paths
                          "Species_chosen", "Surveyx", "breaks"            #settings
                          )))

```

# 1. Area description

## 1.1 Area description - Spatially

The westerschelde is more dynamic (larger flow velocities) and has a salinity gradient, the oosterschelde is less dynamic and no salinity gradient. Both are tidal systems.

The map below shows the emersion-time surface, the land barrier separating the two inlets, and the survey stations coloured by sampling gear.

```{r plot descriptive}


p1 <- ggplot() +
  # Emersion-time raster
  tidyterra::geom_spatraster(data = dvd) +
  scale_fill_viridis_c(
    name    = "Droogvalduur (%)",
    breaks  = breaks,
    na.value = "transparent") +
    scale_fill_binned(palette =  c("#08306B","#08519C","#2171B5","#4292C6","#6BAED6",
                                   "#9ECAE1","#C6DBEF","#DEEBF7","#F7FBFF", 
                                   "lightyellow","lightyellow1","lightyellow2",
                                   "lightyellow3","lightyellow4", "yellow4",NA),
                      breaks = breaks,
                      na.value = "white") +
  # Land
  geom_sf(data = Barrier, fill = "grey85", colour = "grey40", linewidth = 0.2) +
  # Survey points
  geom_sf(data = df_WOT_sf, aes(colour = Tuig), size = 0.6, alpha = 0.6) +
  coord_sf(xlim = c(3.35, 4.30), ylim = c(51.25, 51.75), expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_bw()


 p1 <- p1 + scale_color_manual(values = c("Kokkelschepje"               = "#582f0e",    
                                        "Steekbuis"                     = "#2171B5", # "#bc4749"
                                        "Overig"                        = "#FDE725"))  

print(p1)

```

## 1.2 Area description - Table

```{r Table}

#| message: false

# One row per event, tagged by system (OS east, WS west — they split on longitude).
events_tagged <- df_WOT %>%
  dplyr::distinct(replicate.id, .keep_all = TRUE)

# --- 1. Samples per system per year ---
samples_per_system <- events_tagged %>%
  dplyr::count(Year, Systeem) %>%
  tidyr::pivot_wider(names_from = Systeem, values_from = n, values_fill = 0) %>%
  dplyr::arrange(Year)

knitr::kable(samples_per_system,
             caption = "Number of survey samples per system per year")

# --- 2. Sampled environment (dvd, evmax) per system ---
env_per_system <- events_tagged %>%
  dplyr::filter(!is.na(dvd), !is.na(evmax)) %>%
  dplyr::group_by(Systeem) %>%
  dplyr::summarise(
    n          = dplyr::n(),
    dvd_mean   = mean(dvd),   dvd_sd   = sd(dvd),
    evmax_mean = mean(evmax), evmax_sd = sd(evmax),
    .groups = "drop"
  )

knitr::kable(env_per_system, digits = 2,
             caption = "Sampled emersion time (dvd) and max flow velocity (evmax) by system")

```

# 2. Model

Prepare data for model

```{r Data processing}

events <- df_WOT %>%
  group_by(replicate.id) %>%
  dplyr::summarise(
    Year      = first(Year),
    Longitude = first(Longitude),
    Latitude  = first(Latitude),
    dvd       = first(dvd),
    evmax     = first(evmax),
    Tuig      = first(Tuig),          
    seizoen   = first(seizoen),       
    opp_m2    = first(opp_bemonsterd_m2),
    .groups = "drop"
  )

cockle_counts <- df_WOT %>%
  filter(TaxonName_ACC == Species_chosen) %>%
  group_by(replicate.id) %>%
  dplyr::summarise(count = sum(SumOfNumber, na.rm = TRUE), .groups = "drop")

dat <- events %>%
  left_join(cockle_counts, by = "replicate.id") %>%
  mutate(count = ifelse(is.na(count), 0L, count))

cat("Events with cockles:", sum(dat$count > 0),
    " | true zeros:", sum(dat$count == 0), "\n")

dat <- dat %>%
  filter(!is.na(dvd), !is.na(evmax),
         !is.na(Longitude), !is.na(Latitude), !is.na(opp_m2)) %>%
  mutate(
    dvd_s    = as.numeric(scale(dvd)),     
    evmax_s  = as.numeric(scale(evmax)),   
    log_area = log(opp_m2),                
    gear     = factor(Tuig),
    season   = factor(seizoen),
    fYear    = factor(Year)
  )

## Project lon/lat -> metric CRS (UTM 31N) so mesh & range are in km.
dat <- add_utm_columns(dat, ll_names = c("Longitude", "Latitude"),
                       utm_crs = 32631)   # adds X, Y (km)


```

## 2.1 Constructing a Barrier Mesh

Build a regular SPDE mesh first, then derive a barrier mesh from it by passing the land polygon to `add_barrier_mesh()`. The `range_fraction` shrinks the spatial range across the barrier (lower = stronger barrier), so correlation no longer leaks across land.

```{r mesh}

#first the regular mesh
mesh <- sdmTMB::make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 0.75)
plot(mesh); cat("Mesh vertices:", mesh$mesh$n, "\n")
plot(mesh$mesh, main = "Regular mesh", asp = 1)

Barrier <- sf::st_sf(geometry = Barrier)
#needs tobe mesh units (utm)
Barrier_km <- sf::st_transform(Barrier, 32631)    # metres
sf::st_geometry(Barrier_km) <- sf::st_geometry(Barrier_km) * 1e-3       # -> km to match mesh X/Y
sf::st_crs(Barrier_km) <- NA                      

#Then a Barrier mesh
barrier_mesh <- sdmTMBextra::add_barrier_mesh(
  spde_obj    = mesh,
  barrier_sf  = Barrier_km,       
  range_fraction = 0.2,         # barrier range = 0.2 * estimated range
  proj_scaling = 1,              # ??
  plot = TRUE
)

#plot(barrier_mesh$mesh, main = "Barrier mesh")

```

## 2.2 Fitting the models

```{r Model}

# --- Fit both models 
fit_plain <- sdmTMB(
  count ~ poly(dvd_s, 2) + poly(evmax_s, 2) + gear + fYear,
  data   = dat,
  mesh   = mesh,
  family = nbinom2(link = "log"),
  offset  = dat$log_area,
  spatial = "on"
)

fit_barrier <- sdmTMB(
  count ~ poly(dvd_s, 2) + poly(evmax_s, 2) + gear + fYear,
  data   = dat,
  mesh   = barrier_mesh,    
  family = nbinom2(link = "log"),
  offset  = dat$log_area,
  spatial = "on"
)

```

## 2.3 Setting up Prediction grid

Build the prediction grid from the emersion-time raster (so every prediction cell has a `dvd` value), and predict from each model. I remove subtidal area from prediction grid, as the survey is intertidal.

```{r Prediction-grid}

# Coarsen the emersion raster a little so prediction is quick,

# --- dvd grid
dvd_coarse     <- terra::aggregate(dvd, fact = 4, fun = "mean", na.rm = TRUE)
grid           <- terra::as.data.frame(dvd_coarse, xy = TRUE, na.rm = TRUE)
names(grid)[3] <- "dvd"

# --- pull evmax 
evmax_coarse <- terra::resample(ev_max, dvd_coarse, method = "bilinear")
grid$evmax   <- terra::extract(evmax_coarse,
                               grid[, c("x", "y")])[, 2] 

# drop cells where evmax is missing and filter out subtidal area for prediction grid -> not sure is this is best practice
grid <- grid[!is.na(grid$evmax), ]
grid <- grid[grid$dvd > 0 & grid$dvd < 97, ]

# --- scale BOTH using the fitting means/sds ---
dvd_mean   <- mean(dat$dvd,   na.rm = TRUE); dvd_sd   <- sd(dat$dvd,   na.rm = TRUE)
evmax_mean <- mean(dat$evmax, na.rm = TRUE); evmax_sd <- sd(dat$evmax, na.rm = TRUE)

grid <- grid %>%
  dplyr::rename(Longitude = x, Latitude = y) %>%
  dplyr::mutate(
    dvd_s   = (dvd   - dvd_mean)   / dvd_sd,
    evmax_s = (evmax - evmax_mean) / evmax_sd
  )

# gear still held at reference level 
grid$gear <- factor(levels(dat$gear)[1], levels = levels(dat$gear))
grid$log_area <- 0          

grid$fYear <- factor(2021, levels = levels(dat$fYear))

grid <- add_utm_columns(grid, ll_names = c("Longitude", "Latitude"),
                        utm_crs = 32631)

cat("Prediction cells:", nrow(grid), "\n")

```

# 3. Comparison

## 3.1 Performance

```{r Model Performance}
#| message: false

sanity(fit_plain)
sanity(fit_barrier)

print(fit_plain)
print(fit_barrier)

# AIC 
AIC(fit_plain, fit_barrier)  

#DHARMa residuals  
s1 <- simulate(fit_plain,   nsim = 300, type = "mle-mvn")
s2 <- simulate(fit_barrier, nsim = 300, type = "mle-mvn")

r1 <- dharma_residuals(s1, fit_plain,   return_DHARMa = TRUE)
r2 <- dharma_residuals(s2, fit_barrier, return_DHARMa = TRUE)

DHARMa::testResiduals(r1)
DHARMa::testResiduals(r2)
#DHARMa::testSpatialAutocorrelation(r1, x = dat$X, y = dat$Y)
#DHARMa::testSpatialAutocorrelation(r2, x = dat$X, y = dat$Y)

#S
tidy(fit_plain,   "ran_pars", conf.int = TRUE)
tidy(fit_barrier, "ran_pars", conf.int = TRUE)

```

## 3.2 Prediction

```{r Prediction-predict}

pred_plain   <- predict(fit_plain,   newdata = grid)
pred_barrier <- predict(fit_barrier, newdata = grid)

pred_plain$density   <- exp(pred_plain$est)
pred_barrier$density <- exp(pred_barrier$est)

```

## 3.3 Spatial prediction plots

```{r Prediction-maps, fig.width=11, fig.height=5}

zlim <- range(c(pred_plain$est, pred_barrier$est), na.rm = TRUE)

base_map <- function(df, title) {
  ggplot(df, aes(Longitude, Latitude, fill = est)) +
    geom_sf(data = Barrier, fill = "grey85", colour = "grey40",
            linewidth = 0.2, inherit.aes = FALSE) +
    geom_raster() +
    scale_fill_viridis_c(name = "log expected\ncount",
                         limits = zlim) +             
    coord_sf(xlim = c(3.35, 4.30), ylim = c(51.25, 51.75), expand = FALSE) +
    labs(title = title, x = NULL, y = NULL) +
    theme_bw()
}

p_plain   <- base_map(pred_plain,   "Regular mesh")
p_barrier <- base_map(pred_barrier, "Barrier mesh")

library(patchwork)
p_plain + p_barrier

```

## 

## Reccomendations

Next year (?), the DFS data could be added to explore if mesh/barrier mesh shows a difference in fish species, maybe moving species could react differently?
